---
title: "using package down"
author: "Dale Forrister"
date: "2025-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#routine workflow after edits

Use this whenever you update functions, roxygen docs, README, or vignettes.

Update docs from roxygen

```{r}
devtools::document()       # regenerates NAMESPACE + Rd from #' comments
```

(If README.Rmd changed) knit README
```{r}
devtools::build_readme()   # renders README.Rmd -> README.md
```

(If you added/edited vignettes)

# keep Suggests: knitr, rmarkdown; and VignetteBuilder: knitr in DESCRIPTION
```{r}
devtools::build_vignettes()
```

Quick quality check
```{r}
devtools::check()          # should be clean (0 errors/warnings/notes ideally)
```



Optional: preview site locally

```{r}
pkgdown::clean_site()

pkgdown::build_site()      # view locally; doesn’t affect GitHub
```

Commit & push (source only)

Commit your changes (R/, man/, NAMESPACE, README.md, vignettes/, etc.).

Do not commit built site files if you’re using the GitHub Action (no docs/ or gh-pages changes).

Push to your default branch (e.g., main).

GitHub builds & deploys

GitHub Actions runs pkgdown and updates the gh-pages branch.

Check the Actions tab for a green run, then visit your site URL.

common tasks
```{r}
#Add a vignette

#usethis::use_vignette("intro")  # creates vignettes/intro.Rmd
devtools::build_vignettes()
```

It will appear under the “Articles” menu after the Action rebuilds.

Control homepage

Use README.md at package root for the home page.

Don’t create pkgdown/index.Rmd unless you want to override README.

Customize navigation / reference
Edit _pkgdown.yml (commit it) — examples:

url: https://<user>.github.io/<repo>/
template:
  bootstrap: 5

home:
  title: eCOMET
  description: Tools for ecological metabolomics

reference:
  - title: Core workflow
    contents:
      - GetMZmineFeature
      - AddSiriusAnnot
      - AddChemDist
  - title: Stats & plots
    contents:
      - VolcanoPlot
      - PLSDAplot
      - AnovaBarPlot


News / changelog on the site
```{r}
usethis::use_news_md()

```
Update NEWS.md; it will show under “Changelog”.

Version bump (optional)
```{r}
usethis::use_version("dev")  # or "minor"/"patch" etc.
```
#quick “did something break?” checklist

README not showing? Make sure README.md is in the package root and you don’t have pkgdown/index.*.

404 on site? Confirm Pages is set to gh-pages branch and the Action succeeded.

Vignettes missing? Ensure Suggests contains knitr, rmarkdown; VignetteBuilder: knitr; and that vignettes build locally.

Site not updating? Check the Actions logs for errors, commit _pkgdown.yml if you changed it.

bare-minimum command sequence (copy/paste)
```{r}
devtools::document()
devtools::build_readme()     # if README.Rmd changed
devtools::build_vignettes()  # if vignettes changed
devtools::check()
pkgdown::build_site()  
```
# commit & push; GitHub Action updates the site

That’s it—edit, document, check, push, and let GitHub Actions publish.
