---
title: "eCOMET-intro"
author: "Dale L. Forrister & Min-Soo Choi"
date: "2026-01-26"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eCOMET-intro}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ecomet)
```

## The Mass-spectrometry Metabolomics Object (`mmo`)

The `mmo` is a central data structure in the **eCOMET** package.\
It is created by the `GetMZmineFeature()` function and contains:

-   **feature_data**: A data frame of metabolite features and their
    measured values.
-   **feature_info**: A data frame of feature definitions include mz and
    rt range and grouping information if feature grouping was applied in
    mzmine.
-   **metadata**: Sample metadata, including grouping information.
-   **sirius_annot**: (optional) SIRIUS/CANOPUS class annotations. Added
    by `AddSiriusAnnot()`.
-   **custom_annot**: (optional) Custom compound annotations. Added by
    `AddCustomAnnot()`.
-   **cos.dissim**: (optional) Cosine dissimilarity matrix. Added by
    `AddChemDist()`.
-   **m2ds.dissim**: (optional) MS2DeepScore dissimilarity matrix. Added
    by `AddChemDist()`.
-   **dreams.dissim**: (optional) DreaMS dissimilarity matrix. Added by
    `AddChemDist()`.
-   **\<name\>.dissim**: (optional) Any custom feature dissimilarity
    matrix. Added by `AddCustomDist()` with a user-supplied name.

## Input file requirements

The mendatory input files for creating a `mmo` are:

1.  **MZmine feature table**:

-   This file is generated by MZMine (Feature list methods → Export
    feature list → CSV). See
    [MZMine_documentation](https://mzmine.github.io/mzmine_documentation/module_docs/io/feat-list-export.html)

2.  **Metadata file**: A CSV file containing sample metadata, with
    following columns. -sample (\_\_\_.mzML or equivalent filename used
    as input of the MZmine), -mass (the sample mass which will be used
    for normalization. just put 1 if not needed), -and at least one
    grouping info column (the experimental group for each sample). Use
    as input for 'group_col' argument of `GetMZmineFeature()`.

Refer [Inputs_MZMine](#Inputs_MZMine) below for details.

------------------------------------------------------------------------

You can create a `mmo` like this:

mmo \<- GetMZmineFeature( mzmine_dir = "path/to/mzmine_feature.csv",
metadata_dir = "path/to/metadata.csv", group_col = "treatment" )

Here we load demo data that is downloaded when eCOMET is installed and
create a `mmo`

```{r, eval=FALSE}

# Locate tutorial data shipped with the eCOMET package
data_dir <- system.file(
  "extdata/tutorials/treatment_based",
  package = "ecomet"
)

stopifnot(nzchar(data_dir))  # fail loudly if package/data not installed

# Define file paths
demo_feature <- file.path(data_dir, "feature_table_demo.csv")
demo_metadata <- file.path(data_dir, "metadata_demo.csv")
demo_sirius_formula <- file.path(data_dir, "canopus_formula_summary.tsv")
demo_sirius_structure <- file.path(data_dir, "structure_identifications.tsv")
demo_dreams <- file.path(data_dir, "dreams_sim_demo.csv")
gls_db <- file.path(data_dir, "custom_DB_glucosinolates.csv")


mmo <- GetMZmineFeature(mzmine_dir=demo_feature, metadata_dir = demo_metadata, group_col = 'group', sample_col = 'sample')
```

Once created, you can use the `mmo` for normalization,
annotation, statistical analysis, and visualization throughout the
package.

## Normalization

The raw peak area values are processed by following steps:

1.  Replacing missing values (NA) with half of the minimum value for
    each feature or 1.

2.  Normalizing by sample mass (provided in metadata).

3.  Transform the data using Log, Mean-centering, and Z-score scaling.

```{r, eval=FALSE}
mmo <- ReplaceZero(mmo, method = 'one') # Replace 0 and NA values by 1
mmo <- MassNormalization(mmo) # Normalize peak area by sample mass in metadata
mmo <- MeancenterNormalization(mmo) # Add mean-centered area
mmo <- LogNormalization(mmo) # Add log-transformed area
mmo <- ZNormalization(mmo) # Add Zscore
```

## Adding Sirius/CANOPUS annotations

The results from SIRIUS/CANOPUS can be added to the `mmo` as
**mmo\$sirius_annot** These are used for chemical class enrichment
analysis and visualization. Two files are needed:

1.  **structure_identifications.tsv**

2.  **canopus_formula_summary.tsv**

The above files are generated by SIRIUS/CANOPUS. See [SIRIUS
documentation](https://v6.docs.sirius-ms.io/) and the
[Inputs_SIRIUS](#Inputs_SIRIUS) section at the bottom of this page for
details.

mmo \<- AddSiriusAnnot( mmo, structure_dir =
"path/to/structure_identifications.tsv", canopus_dir =
"path/to/canopus_formula_summary.tsv" )

```{r, eval=FALSE}
mmo <- AddSiriusAnnot(mmo, 
                      canopus_structuredir = demo_sirius_structure, 
                      canopus_formuladir = demo_sirius_formula)
mmo$sirius_annot
```

## Adding chemical dissimilarity

Chemical dissimilarity matrices describe pairwise structural
relationships among features. Once added to the `mmo` they are
used by `GetBetaDiversity()`, `GetAlphaDiversity()` (weighted mode), and
`HeatmapPlot()` via the `distance` argument.

### Built-in methods: `AddChemDist()`

`AddChemDist()` reads MZmine molecular networking output files for three
supported similarity types and converts them to dissimilarity
(`1 - similarity`) before storing them in the `mmo`.

``` r
mmo <- AddChemDist(mmo,
  cos_dir    = "path/to/cosine_similarity.csv",
  m2ds_dir   = "path/to/ms2deepscore_similarity.csv",
  dreams_dir = "path/to/dreams_similarity.csv"
)
```

```{r, eval=FALSE}
# Add Dreams distance from demo data
mmo <- AddChemDist(mmo, dreams_dir = demo_dreams)
mmo$dreams.dissim[1:10, 1:10]
```

Pass the stored matrix to downstream functions using its short name:

```{r, eval=FALSE}
beta <- GetBetaDiversity(mmo, method = "CSCS", distance = "dreams")
alpha <- GetAlphaDiversity(mmo, mode = "weighted", distance = "dreams")
```

### Custom distance matrices: `AddCustomDist()`

If you have a feature dissimilarity matrix from another tool (e.g.
Tanimoto structural similarity, NPC-based distances, or any other
pairwise metric), use `AddCustomDist()` to add it under a name of your
choice.

**Format requirements for the matrix:**

-   Square numeric matrix with row and column names matching the feature
    IDs in `mmo$feature_data$id`.
-   Values must be **dissimilarities** in `[0, 1]`: 0 = identical, 1 =
    maximally different. If your source is a similarity matrix, convert
    it first with `dist_matrix <- 1 - sim_matrix`.
-   Diagonal should be 0. Matrix should be symmetric.

```{r, eval=FALSE}
# Example: Tanimoto structural similarity from an external tool
tanimoto_dist <- 1 - tanimoto_sim_matrix  # convert similarity → dissimilarity

mmo <- AddCustomDist(mmo, dist_matrix = tanimoto_dist, name = "tanimoto")

# Use it anywhere a 'distance' argument is accepted:
beta  <- GetBetaDiversity(mmo, method = "CSCS",     distance = "tanimoto")
alpha <- GetAlphaDiversity(mmo, mode = "weighted",  distance = "tanimoto")
```

The matrix is stored as `mmo$tanimoto.dissim` and the name `"tanimoto"`
can be passed as the `distance` argument throughout the package.

## Adding custom annotations

You can add your own custom compound annotations to the `mmo` as
**mmo\$custom_annot**. This is useful for adding known compound names or
classes from other databases. The custom annotation file should be a CSV
with three columns: 1. compound: Name of the compound 2. mz: m/z value
3. rt: Retention time (in minutes)

```{r, eval=FALSE}
mmo <- AddCustomAnnot(
  mmo,
  custom_annot_dir = "path/to/custom_annotations.csv",
  mztol = 5, # m/z tolerance in ppm
  rttol = 0.2 # RT tolerance in minutes
)
```

##Filtering `mmo`s based on feature list or sample list

You can provide either a list of species or features to filter the
`mmo`. This will iterate over all sets of

```{r, eval=FALSE}

#Create a new mmo containing only control samples
samples_ctrl <- mmo$metadata$sample[mmo$metadata$group == "ctrl"]

mmo_ctrl <- filter_mmo(mmo,
                       sample_list  = samples_ctrl,
                       drop_empty_feat = TRUE #This will remove any features not observed in samples
                       )


```

```{r, eval=FALSE}

#create a new mmo only containing Flavonoid based on sirius annotations
FLVs <- mmo$sirius_annot %>% 
  filter(grepl("Flavonoid", .data[['ClassyFire#most specific class']])
  ) %>% pull(id)

mmo_FLVs <- filter_mmo(mmo,
                       feature_list = FLVs)

```

##Match and Filter an associated mgf file to based on the feature list
in a `mmo`.

Once an dataset has been filtered you may want to update the associated
.mgf file containing ms2 spectra prior to processing with Sirius, GNPS
ect.

1)  Match spectra present in an MGF against a feature list and update
    the mmo\$feature_info.

```{r, eval=FALSE}

mmo <- annotate_feature_info_ms2_from_mgf(mmo,
                                          mgf_path = "/path_to_mgf/spectra.mgf",
                                          overwrite = TRUE)
mmo$feature_info
```

2)  Given a feature list in a `mmo` defined by `mmo$feature_data$id`
    read through an associated mgf file and remove features not in the
    list.

```{r, eval=FALSE}
filter_mgf_to_mmo(mmo, 
                  mgf_path = "/path_to_mgf/spectra.mgf",
                  output_path = "/path_to_renamed_mgf/filtered_spectra_mmo_only.mgf")
```

------------------------------------------------------------------------
