---
title: "Interspecific Ecometabolomics Tutorial"
output: html_notebook
---

This Relative Paths will work if you open the .Rjroject. 
```{r, eval = FALSE}
library(here)
here()

library(ecomet)
library(dplyr)
```
#1) Created the mmo object
```{r, eval = FALSE}

demo_feature <- here("tutorial/tutorial_2_interspecific/Ecomet_Interspecific_Demo_full_feature_table.csv")
demo_metadata <- here("tutorial/tutorial_2_interspecific/Ecomet_Interspecific_Demo_metadata_no_blanks.csv")
demo_sirius_formula <- here("tutorial/tutorial_2_interspecific/sirius/formula_identifications.tsv")
demo_sirius_structure <- here("tutorial/tutorial_2_interspecific/sirius/structure_identifications.tsv")
demo_dreams <- here("tutorial/tutorial_2_interspecific/Ecomet_Interspecific_Demo_dreams_sim_dreams.csv")


# Initialize eCOMET object
mmo <- GetMZmineFeature(mzmine_dir=demo_feature, metadata_dir = demo_metadata, group_col = 'Species',sample_col = "filename")
mmo <- AddFeatureInfo(mmo = mmo,full_feature_csv=demo_feature)
# Add normalizations
mmo <- ReplaceZero(mmo, method = 'one') # Replace 0 and NA values by 1
mmo <- LogNormalization(mmo) # Add log-transformed area
mmo <- ZNormalization(mmo) # Add Zscore



```
Now the mmo object contains the Feature x Abundance matrix where the first two columns identify individual metabolites and the remaining columns represent the abundance in each sample
```{r, eval = FALSE}
mmo$feature_data
```
We can add the feature x feature pairwise distance to the mmo object based on the output of MzMine. This can accept modified cosine, dreaMS and MS2Deepscore
```{r, eval = FALSE}
# Add Dreams distance
mmo <- AddChemDist(mmo, dreams_dir = demo_dreams)

mmo$dreams.dissim[1:10,1:10]
```

Sample Metadata:
In this tutorial we are comparing 5 species of hickory along side a sample of Eastern black walnut.

mmo$metadata$Species

There are also two types of samples "pools" and "samples. Pools represent species pools where each individual sample for each spcies was combined and run for ms/ms data analysis. Samples represent MS1 data for anindividual plant.


```{r, eval = FALSE}
mmo$metadata
```
Note we can subset an mmo object using filter mmo to filte based on a list of samples, a list of groups or a list of features

In this tutorial we will subset to only individual samples removing the pooled samples
```{r, eval = FALSE}

samples_keep <- mmo$metadata$sample[mmo$metadata$Type != "pool"]
mmo_individual <- filter_mmo(mmo,sample_list  = samples_keep,drop_empty_feat = TRUE)
```


# 2) Visualize distance between samples with PLSDA

```{r, eval = FALSE}

colors <- c(
  "Carya cordiformis" = "#4C72B0",  # blue
  "Carya glabra"      = "#55A868",  # green
  "Carya ovata"       = "#C44E52",  # red
  "Carya tomentosa"   = "#8172B2",  # purple
  "Juglans nigra"     = "#CCB974"   # amber/gold
)



PCAplot(mmo = mmo_individual,color = colors,outdir = "~/Downloads/test_pca.pdf",label = T,normalization = 'None')
PLSDAplot(mmo = mmo_individual,color = colors,outdir = "~/Downloads/test_pca.pdf",normalization = 'None')


```
# 3) Calculate Beta Diversity between samples
```{r, eval = FALSE}
#First we calculate dreams
beta_diversity_dreams <- GetBetaDiversity(mmo_individual, method = 'CSCS', 
 normalization = 'None', distance = 'dreams', filter_feature = FALSE)

```

