---
title: "Intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette-eCOMET-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ecomet)
```

## The `mmo` object

The `mmo` object is a central data structure in the **eCOMET** package.  
It is created by the `GetMZmineFeature()` function and contains:

- **feature_data**: A data frame of metabolite features and their measured values.
- **metadata**: Sample metadata, including grouping information.
- **pairwise**: (optional) Results of pairwise comparisons between features. Added by `PairwiseComp()`.
- **sirius_annot**: (optional) SIRIUS/CANOPUS class annotations. Added by `AddSiriusAnnot()`.
- **custom_annot**: (optional) Custom compound annotations. Added by `AddCustomAnnot()`.
- **cos.dissim**: (optional) Cosine dissimilarity matrix. Added by `AddChemdist()`.
- **m2ds.dissim**: (optional) MS2DeepScore dissimilarity matrix. Added by `AddChemdist()`.
- **dreams.dissim**: (optional) DreaMS dissimilarity matrix. Added by `AddChemdist()`.

You can create an `mmo` object like this:

```{r, eval = FALSE}
mmo <- GetMZmineFeature(
  mzmine_dir = "path/to/mzmine_feature.csv",
  metadata_dir = "path/to/metadata.csv",
  group_col = "treatment"
)
```

Once created, you can use the `mmo` object for normalization, annotation, statistical analysis, and visualization throughout the package.

## Input file requirements
The mendatory input files for creating an `mmo` object are:
1. **MZmine feature table**: 
  - This file is generated by MZMine (Feature list methods → Export feature list → CSV). See [MZMine_documentation](https://mzmine.github.io/mzmine_documentation/module_docs/io/feat-list-export.html) 
2. **Metadata file**: A CSV file containing sample metadata, with following columns. See [Metadata_example.csv]()
  -sample (___.mzML or equivalent filename used as input of the MZmine), 
  -mass (the sample mass which will be used for normalization. just put 1 if not needed), 
  -and at least one grouping info column (the experimental group for each sample). Use as input for 'group_col' argument of `GetMZmineFeature()`.
Refer [Inputs_MZMine](#Inputs_MZMine) below for details.
## Normalization
The raw peak area values are processed by following steps:
1. Replacing missing values (NA) with half of the minimum value for each feature or 1.
2. Normalizing by sample mass (provided in metadata).
3. Transform the data using Log, Mean-centering, and Z-score scaling.

```{r, eval = FALSE}
mmo <- ReplaceZero(mmo, method = 'one') # Replace 0 and NA values by 1
mmo <- MassNormalization(mmo) # Normalize peak area by sample mass in metadata
mmo <- MeancenterNormalization(mmo) # Add mean-centered area
mmo <- LogNormalization(mmo) # Add log-transformed area
mmo <- ZNormalization(mmo) # Add Zscore
```

## Adding Sirius/CANOPUS annotations
The results from SIRIUS/CANOPUS can be added to the `mmo` object as **mmo$sirius_annot**
These are used for chemical class enrichment analysis and visualization.
Two files are needed:
1. **structure_identifications.tsv**
2. **canopus_formula_summary.tsv**
The above files are generated by SIRIUS/CANOPUS. See [SIRIUS documentation](https://v6.docs.sirius-ms.io/) and the [Inputs_SIRIUS](#Inputs_SIRIUS) section at the bottom of this page for details.
```{r, eval = FALSE}
mmo <- AddSiriusAnnot(
  mmo,
  structure_dir = "path/to/structure_identifications.tsv",
  canopus_dir = "path/to/canopus_formula_summary.tsv"
)
```

## Adding chemical dissimilarity
The chemical dissimilarity matrices (cosine, MS2DeepScore, DreaMS) can be added to the `mmo` object.
These are used for chemical dendrograms and chemical diversity analyses.
```{r, eval = FALSE}
mmo <- AddChemdist(
  mmo,
  cos_dir = "path/to/cosine_dissimilarity.csv",
  m2ds_dir = "path/to/ms2deepscore_dissimilarity.csv",
  dreams_dir = "path/to/dreams_dissimilarity.csv"
)
```

## Adding custom annotations
You can add your own custom compound annotations to the `mmo` object as **mmo$custom_annot**.
This is useful for adding known compound names or classes from other databases.
The custom annotation file should be a CSV with three columns:
1. compound: Name of the compound
2. mz: m/z value
3. rt: Retention time (in minutes)
```{r, eval = FALSE}
mmo <- AddCustomAnnot(
  mmo,
  custom_annot_dir = "path/to/custom_annotations.csv",
  mztol = 5, # m/z tolerance in ppm
  rttol = 0.2 # RT tolerance in minutes
)
```

# Preparing input files
## Inputs_MZMine

## Inputs_SIRIUS