---
title: "Treatment-based_study_tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Treatment-based_study_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# pak::pak("phytoecia/eCOMET")
library(ecomet)
library(dplyr)
library(stringr)
```

## Background
In this tutorial, we will demonstrate how to use the **eCOMET** package for analyzing metabolomics data from a treatment-based study.
- Treatment-based studies involve comparing metabolite profiles between different treatment groups (e.g., control vs treated).
- The tutorial covers data preprocessing, normalization, statistical analysis, and visualization techniques commonly used in treatment-based metabolomics studies.
- The tutorial files are metabolomics analysis files of *Arabidopsis thaliana* (Col-0) attacked by two different herbivores (*Spodoptera litura*; sl, and *Lipaphis erysimi*; le). 
- Eight replicates were sampled and analyzed for each group. See metadata.csv for more details. 
- Make sure you downloaded the tutorial files from [here]() and set the working directory to the downloaded folder.

```{r, eval = FALSE}
setwd('/path/to/eCOMET/tutorial/tutorial_1_treatment_based')

demo_feature <- "raw_data/feature_table_demo.csv"
demo_metadata <- "raw_data/metadata_demo.csv"
demo_sirius_formula <- "raw_data/canopus_formula_summary.tsv"
demo_sirius_structure <- "raw_data/structure_identifications.tsv"
demo_dreams <- "raw_data/dreams_sim_demo.csv"
gls_db <- "raw_data/custom_DB_glucosinolates.csv"
```

## 1. Create an `mmo` object
Following steps are performed to create an `mmo` object and add various normalizations and annotations. Inspect the structure of the 'mmo' object using 'summary(mmo)'' after each step to see how the object is updated.
```{r, eval = FALSE}
# Initialize eCOMET object
mmo <- GetMZmineFeature(mzmine_dir=demo_feature, metadata_dir = demo_metadata, group_col = 'group')
# Add normalizations
mmo <- ReplaceZero(mmo, method = 'one') # Replace 0 and NA values by 1
mmo <- MassNormalization(mmo) # Normalize peak area by sample mass in metadata
mmo <- MeancenterNormalization(mmo) # Add mean-centered area
mmo <- LogNormalization(mmo) # Add log-transformed area
mmo <- ZNormalization(mmo) # Add Zscore

# Add SIRIUS annotation
mmo <- AddSiriusAnnot(mmo, canopus_structuredir = demo_sirius_structure, canopus_formuladir = demo_sirius_formula)
# Make a vector of flavonoids using SIRIUS annotation
FLVs <- mmo$sirius_annot %>% filter(str_detect(mmo$sirius_annot[['ClassyFire#most specific class']], "Flavonoid")) %>% pull(feature)

# Add custom annotation using inhouse glucosinolate library
mmo <- AddCustomAnnot(mmo, DB = gls_db, mztol = 5, rttol = 0.2)
# Make a vector of glucosinolates using custom annotation
GLSs <- mmo$custom_annot %>% filter(lengths(custom_annot) > 0) %>% pull(feature)

# Add Dreams distance
mmo <- AddChemDist(mmo, dreams_dir = demo_dreams)
```
## 2. Plot dimensionality reduction plots
### PCA plot and PERMANOVA
A PCA plot is commonly used to visualize the overall distribution of groups.
Following PERMANOVA test can be performed to check if the groups are significantly different.

```{r, eval = FALSE}
# Set colors for groups
colors <- c("ctrl" = "grey", "sl1" = "#fdcdac", "le1" = "#b3e2cd")
# PCA plot
# A plot and permanova output files will be generated in the specified outdir
PCAplot(mmo, color = colors, outdir = 'plot/PCA')
PCAplot(mmo, color = colors, outdir = 'plot/PCA_log', label = FALSE, normalization = 'Log') # Data normalization option
PCAplot(mmo, color = colors, outdir = 'plot/PCA_FLV', label = FALSE, filter_feature = TRUE, feature_list = FLVs) # Feature filtering option
```
### PLS-DA plot
```{r, eval = FALSE}
# Plot PLS-DA
PLSDAplot(mmo, color = colors, outdir = 'plot/PLSDA.pdf')
PLSDAplot(mmo, color = colors, outdir = 'plot/PLSDA_meancentered.pdf', normalization = 'Meancentered') # Data normalization option
```

## 3. Identify differentially accumulated metabolites (DAMs)
- 'PairwiseComp()' function performs pairwise comparison between two groups using t-test and fold change. The results are stored in mmo$pairwise_comp.
- 'GetDAMs()' function extracts DAMs based on user-defined cutoffs for p-value and fold change.
- Here, we compare each herbivore treatment group (sl1 and le1) to the control group (ctrl).
- Log2(1.5) = 0.5849625 is used as fold change cutoff and 0.1 as p-value cutoff to extract more DAMs
```{r, eval = FALSE}
# Run pairwise comparison
mmo <- PairwiseComp(mmo, group1 = 'ctrl', group2 = 'sl1')
mmo <- PairwiseComp(mmo, group1 = 'ctrl', group2 = 'le1')
# Extract DAMs
DAMs <- GetDAMs(mmo, fc_cutoff = 0.5849625, pval_cutoff = 0.1) # log2(1.5) = 0.5849625
DAMs_up <- DAMs$DAMs_up
DAMs_down <- DAMs$DAMs_down
head(DAMs_up)
```

### 3.1. Volcano plot
A volcano plot can be used to visualize the overall distribution of features based on fold change and p-value.
```{r, eval = FALSE}
# Volcano plot
VolcanoPlot(mmo, comp = 'ctrl_vs_sl1', outdir = 'plot/Volcano_ctrl_vs_sl1.pdf')
VolcanoPlot(mmo, comp = 'ctrl_vs_le1', outdir = 'plot/Volcano_ctrl_vs_le1.pdf', topk = 0) # Remove label by setting topk = 0
```
### 3.2. Venn diagram and upset plot
A Venn diagram or upset plot can be used to visualize the overlap of DAMs between different comparisons.
```{r, eval = FALSE}
# Define input list
VennInput <- list(
  sl1.up = DAMs_up$ctrl_vs_sl1.up,
  le1.up = DAMs_up$ctrl_vs_le1.up
)
# Venn diagram
require(ggvenn)
library(ggvenn)
ggvenn(VennInput, stroke_size = 0.5, set_name_size = 4, show_percentage = FALSE) +
  theme(legend.position = "none") 
ggsave("plot/Venn_Upreg.pdf", height = 5, width = 5)
# UpSet plot
require(UpSetR)
library(UpSetR)
pdf("plot/Upset_Upreg.pdf", 7, 5)
upset(fromList(VennInput), nsets=10, nintersects=20,order.by='freq', mainbar.y.label='Features in Set', line.size=1, point.size=4, shade.color='white', text.scale=1, show.numbers=FALSE)
dev.off()
```

## 4. Heatmap 
To visualize the relative abundance of features across samples, a heatmap can be generated.
The features can be either clusterd using hierarchical clustering or by chemical distances (e.g., Dreams distance), 
following idea of Qemistree (Tripathi et al., 2021, Nat Chem. Biol.).

- The input matrix for heatmap can be either fold change between each treatment and control or mean normalized values across samples.
- Various normalizations can be used (None, Log, Meancentered, and Z).
- Groups or features can be filtered by setting 'filter_group' or 'filter_feature' to TRUE and providing a list of groups or features.
- The input data for the heatmap is generated by 'GenerateHeatmapInputs()' function, then the heatmap is plotted using 'pheatmap' package.
```{r, eval = FALSE}
library(pheatmap)
# Generate input for heatmap
# distance is one of the chemical distance (cosine, m2ds, and dreams) for clustering rows
# The values can be either fold_change or mean (use option 'summarize')
heatmap_inputs <- GenerateHeatmapInputs(
  mmo, summarize = 'fold_change', control_group = 'ctrl', 
  normalization = 'None', distance = 'dreams'
) # This generates fold change matrix between each treatment and control

# The resulting list contains FC_matrix, dist_matrix, row_label, and heatmap_data
# A heatmap can be generated using pheatmap
pdf("plot/Heatmap_FC_dreams.pdf", width = 10, height = 10)
pheatmap(mat = heatmap_inputs$FC_matrix, 
     clustering_distance_rows = heatmap_inputs$dist_matrix,  
     clustering_method = "average", 
     cellwidth = 100,
     cellheight = 0.3,
     treeheight_row = 100,
     fontsize_row = 3,
     fontsize_col = 15,
     scale = 'none'
)
dev.off()

# Either, you can visualize mean normalized values across samples
heatmap_inputs <- GenerateHeatmapInputs(
  mmo, summarize = 'mean', normalization = 'Z', distance = 'dreams'
) 
# 'clustering_distance_rows' option make the dendrogram follows chemical distances of features. 
#  -Delete this option to visualize the heatmap following cannonical clustering
pdf("plot/Heatmap_Mean_Z_clustering.pdf", width = 10, height = 10)
pheatmap(mat = heatmap_inputs$FC_matrix, 
     #clustering_distance_rows = heatmap_inputs$dist_matrix,  # Delete this option to visualize the heatmap following cannonical clustering 
     clustering_method = "average", #UPGMA
     cellwidth = 100,
     cellheight = 0.3,
     treeheight_row = 100,
     fontsize_row = 3,
     fontsize_col = 15,
     scale = 'none'
)
dev.off()

# Feature filtering option
# Visualize only glucosinolates
# As glucosinolate annotations are stored in mmo$custom_annot (by AddCustomAnnot() function),
# the compound names are stored in heatmap_inputs$row_label and can be used as row names
heatmap_inputs_GLS <- GenerateHeatmapInputs(
  mmo, summarize = 'mean', normalization = 'Z', distance = 'dreams',
  filter_feature = TRUE, feature_list = GLSs
) 
pdf("plot/Heatmap_Mean_Z_GLS.pdf", width = 10, height = 10)
pheatmap(mat = heatmap_inputs_GLS$FC_matrix, 
     #clustering_distance_rows = heatmap_inputs_GLS$dist_matrix,  # Delete this option to visualize the heatmap following cannonical clustering 
     clustering_method = "average", #UPGMA
     cellwidth = 100,
     cellheight = 8,
     treeheight_row = 100,
     fontsize_row = 8,
     fontsize_col = 15,
     scale = 'none',
     annotation_names_row = TRUE,
     labels_row = heatmap_inputs_GLS$row_label
)
dev.off()
```

## 5. CANOPUS class enrichment analysis
To identify overepresented chemical classes in a set of features of interest (e.g., DAMs), CANOPUS class enrichment analysis can be performed following idea of ORA.

```{r, eval = FALSE}
# For a single set of features, a detailed enrichment plot can be generated
# There are two plotting styles available
CanopusListEnrichmentPlot(mmo, DAMs_up$ctrl_vs_sl1.up, pthr = 0.1, outdir = 'plot/sl1_up_enrichment.pdf', height = 6, width = 6)
CanopusListEnrichmentPlot_2(mmo, DAMs_up$ctrl_vs_sl1.up, pthr = 0.1, outdir = 'plot/sl1_up_enrichment_2.pdf', topn = 10, height = 6, width = 6)

# For a list of sets features, a summary enrichment plot can be generated
# The summary enrichment plot can be generated for either a single level of CANOPUS classification (8.2.1) or for all levels (8.2.2)
# For a single level of CANOPUS classification
term_levels <- c('NPC_class', 'NPC_superclass', 'NPC_pathway', 'ClassyFire_most_specific', 'ClassyFire_level5', 'ClassyFire_subclass', 'ClassyFire_class', 'ClassyFire_superclass')
CanopusLevelEnrichmentPlot(mmo, DAMs_up, term_level = 'NPC_class', pthr = 0.1, prefix = 'plot/DAMs_up_NPC_class')


# For all levels of CANOPUS classification
# All levels, or only NPC or ClassyFire
CanopusAllLevelEnrichmentPlot(mmo, DAMs_up, term_level = 'NPC', pthr = 0.1, prefix = 'plot/DAMs_up_all_terms', width = 8, height = 12)
CanopusAllLevelEnrichmentPlot(mmo, DAMs_up, term_level = 'ClassyFire', pthr = 0.1, prefix = 'plot/DAMs_up_all_terms', width = 8, height = 12)
```
