---
title: "vignette-eCOMET-intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette-eCOMET-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ecomet)
```

## `mmo` object?

The `mmo` object is a central data structure in the **eCOMET** package.  
It is created by the `GetMZmineFeature()` function and contains:

- **feature_data**: A data frame of metabolite features and their measured values.
- **metadata**: Sample metadata, including grouping information.
- **pairwise**: Results of pairwise comparisons between features. Added by `PairwiseComp()`.
- **sirius_annot**: (optional) SIRIUS/Canopus chemical annotations. Added by `AddSiriusAnnot()`.
- **custom_annot**: (optional) Custom compound annotations. Added by `AddCustomAnnot()`.

You can create an `mmo` object like this:

```{r, eval = FALSE}
mmo <- GetMZmineFeature(
  mzmine_dir = "path/to/mzmine_feature.csv",
  metadata_dir = "path/to/metadata.csv",
  group_col = "treatment"
)
```

Once created, you can use the `mmo` object for normalization, annotation, statistical analysis, and visualization throughout the package.

## Input file requirements
The mendatory input files for creating an `mmo` object are:
1. **MZmine feature table**: 
2. **Metadata file**: A CSV file containing sample metadata, with following columns:
  -sample (___.mzML or equivalent filename used as input of the MZmine), 
  -mass (the sample mass which will be used for normalization. just put 1 if not needed), 
  -and at least one grouping info column (the experimental group for each sample).
3. 

See the demo files for examples.
### MZmine feature table
Exported from MZmine as a CSV file. 
### Metadata file
A CSV file with at least three columns: `sample`, `mass`, and one grouping column (e.g., `treatment`).
The `sample` column should match the filenames in the MZmine feature table. 
The `mass` column is used for normalization (set to 1 if not needed). 
The grouping column is used for statistical comparisons.